---
title: "hw5_erhs535_erh"
author: "Eliud R. Rivas Hernandez"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

# Introduction   

This analysis examines the spatial distribution of homicides in Baltimore, MD using data from the Washington Post's comprehensive investigation of unsolved homicides in major U.S. cities. The analysis uses the `sf` framework for spatial analysis and `tigris` to provide geographic context through census tract boundaries.

## Questions of Interest:
1. How are homicides spatially distributed across Baltimore?  

2. Are there geographic differences between solved and unsolved homicides?    
3. Do different racial groups experience homicides in different geographic areas? 

# Load required packages:

```{r packages}

library(tidyverse)
library(sf)
library(tigris)
library(forcats)
library(viridis)
library(scales)

```

# Data Loading and Preparation

## Read Homicide Data

```{r load-data}

# Read the Washington Post homicide data
homicides <- read_csv("data/homicide-data.csv")

# Display structure
glimpse(homicides)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
